---
title: "lb_project"
output: html_document
date: "2026-07-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LB Project

This project is intended to help you apply your skills in R in a way that would be conducive for collaborating or creating reproducible analyses.


## Data Used

Identify a dataset to analyze. Remember that if you share the data you must **not violate any privacy restrictions**

If you don't have your own data, options for places to find data are listed on https://daseh.org.

You might start by setting up an R Project, importing your data, previewing the data, etc. 

```{r}
library(tidyverse)
library(broom)
library(naniar)
library(maps)
library(usmap)
library(tidycensus)
library(scales)
dat <- read_csv("https://drive.usercontent.google.com/download?id=1V_1-6dfl88E6R_F3NbedqUCjmFabyMJ7&export=download&authuser=0&confirm=t&uuid=9b3505cb-8033-4650-be3e-24d2e914609c&at=AAINaIJgGDkjG47-KSbtWO1UTWTq:1782243211418")
dat
```

**describe** what your data looks like and **where** you got it. Identify what the variables and samples are, or point to a data dictionary. Describe how the data was originally created. 

US census data for Long Beach, California. Wide data with 26 columns. I got this cleaned US census data for Long Beach from Data Day 2026 at CSULB. Samples are surveyed individuals and the variables are; 
SERIALNO
Pwgtp 
Year
age
Sex
Location
Race_eth
Age_cat
Education
Personal_income
Household_income
Class
Has_medicaid
Has_health_insurance
Has_disability
Has_food_stamps
Tenure
Income_to_poverty_ratio
Below_poverty_line
Household
Hisp_Ethnicity
AAPI_Ethnicity

```{r}
nrow(dat)
ncol(dat)
colnames(dat)
miss_var_which(dat)
```


# Data Cleaning/Wrangling

Perform any data subsetting, cleaning, or manipulation methods that were described in this course on your data. Examples are: renaming the columns, recoding values, reshaping the data, filtering the data etc. 

You might also research other methods to include here, especially if your data is very unique.
```{r}
#data subset for 2023, exclude 2021
dat |> pull(year) |> range()
dat |> filter(year == 2021) |> nrow()
dat |> filter(year == 2022) |> nrow()
dat |> filter(year == 2023) |> nrow()
subdat <- dat |> filter(year == 2023)

subdat |> pull("has_medicaid") |> class()
#turn character columns into numeric to calculate percentages later
subdat <- subdat |> mutate(has_medicaid = if_else(has_medicaid == "Yes", 1L, 0L),
has_health_insurance = if_else(has_health_insurance == "Yes", 1L, 0L),
has_disability = if_else(has_disability == "Yes", 1L, 0L),
has_food_stamps = if_else(has_food_stamps == "Yes", 1L, 0L)
)

#public assistance column has lots of 0 and -1 which will mess with averages later, I'll replace with na and omit during calculations 
#In U.S. Census Bureau microdata a -1 is a missing data flag or a placeholder.
subdat <- subdat |> replace_with_na(replace = list(public_assistance_income = c(-1,0)))
subdat
```





# Data Analysis

Perform a **simple analysis** of your data. This can involve summarizing the data to describe aspects about it (quartiles, means, range etc.) or a simple statistical test. 

Don't forget to describe what analysis you performed and why. Provide some simple **interpretation** about what your analysis might indicate about your data.

```{r}
dat |> pull(location) |> unique() 
#total amount surveyed individuals per lb neighborhood
east_total <- subdat |> filter(location == "East Long Beach") |> nrow()
south_total <- subdat |> filter(location == "South Long Beach") |> nrow()
north_total <- subdat |> filter(location == "North Long Beach") |> nrow()
central_total <- subdat |> filter(location == "Central Long Beach") |> nrow()

#calculate percentage of surveyed individuals on public assistance (medicaid, disability, food stamps) or have health insurance

#health insurance
health_e <- subdat |> filter(location == "East Long Beach") |> pull(has_health_insurance) |> sum() / east_total 
health_s <- subdat |> filter(location == "South Long Beach") |> pull(has_health_insurance) |> sum() / south_total
health_n <- subdat |> filter(location == "North Long Beach") |> pull(has_health_insurance) |> sum() / north_total
health_c <- subdat |> filter(location == "Central Long Beach") |> pull(has_health_insurance) |> sum() / central_total

percentage <- function(x) x * 100
health_e <- percentage(x = health_e)
health_s <- percentage(x = health_s)
health_n <- percentage(x = health_n)
health_c <- percentage(x = health_c)

#medicaid
med_e <- subdat |> filter(location == "East Long Beach") |> pull(has_medicaid) |> sum() / east_total 
med_s <- subdat |> filter(location == "South Long Beach") |> pull(has_medicaid) |> sum() / south_total
med_n <- subdat |> filter(location == "North Long Beach") |> pull(has_medicaid) |> sum() / north_total
med_c <- subdat |> filter(location == "Central Long Beach") |> pull(has_medicaid) |> sum() / central_total

med_e <- percentage(x = med_e)
med_s <- percentage(x = med_s)
med_n <- percentage(x = med_n)
med_c <- percentage(x = med_c)

#disability
able_e <- subdat |> filter(location == "East Long Beach") |> pull(has_disability) |> sum() / east_total 
able_s <- subdat |> filter(location == "South Long Beach") |> pull(has_disability) |> sum() / south_total
able_n <- subdat |> filter(location == "North Long Beach") |> pull(has_disability) |> sum() / north_total
able_c <- subdat |> filter(location == "Central Long Beach") |> pull(has_disability) |> sum() / central_total

able_e <- percentage(x = able_e)
able_s <- percentage(x = able_s)
able_n <- percentage(x = able_n)
able_c <- percentage(x = able_c)

#food stamps
food_e <- subdat |> filter(location == "East Long Beach") |> pull(has_food_stamps) |> sum() / east_total 
food_s <- subdat |> filter(location == "South Long Beach") |> pull(has_food_stamps) |> sum() / south_total
food_n <- subdat |> filter(location == "North Long Beach") |> pull(has_food_stamps) |> sum() / north_total
food_c <- subdat |> filter(location == "Central Long Beach") |> pull(has_food_stamps) |> sum() / central_total

food_e <- percentage(x = food_e)
food_s <- percentage(x = food_s)
food_n <- percentage(x = food_n)
food_c <- percentage(x = food_c)

#average personal income per neighborhood
money_e <- subdat |> filter(location == "East Long Beach") |> pull(personal_income) |> mean()
money_s <- subdat |> filter(location == "South Long Beach") |> pull(personal_income) |> mean()
money_n <- subdat |> filter(location == "North Long Beach") |> pull(personal_income) |> mean()
money_c <- subdat |> filter(location == "Central Long Beach") |> pull(personal_income) |> mean()

#average household income per neighborhood
house_e <- subdat |> filter(location == "East Long Beach") |> pull(household_income) |> mean()
house_s <- subdat |> filter(location == "South Long Beach") |> pull(household_income) |> mean()
house_n <- subdat |> filter(location == "North Long Beach") |> pull(household_income) |> mean()
house_c <- subdat |> filter(location == "Central Long Beach") |> pull(household_income) |> mean()


#how many people have public assistance income in long beach total: 80
public <- subdat |>  na.omit() |> nrow()

#percentage people have public assistance income in long beach per neighborhood
avg_pub_e <- subdat |> filter(location == "East Long Beach") |> na.omit() |> nrow() / east_total
avg_pub_e <- percentage(x = avg_pub_e)

avg_pub_s <- subdat |> filter(location == "South Long Beach") |> na.omit() |> nrow() / south_total
avg_pub_s <- percentage(x = avg_pub_s)

avg_pub_n <- subdat |> filter(location == "North Long Beach") |> na.omit() |> nrow() / north_total
avg_pub_n <- percentage(x = avg_pub_n)

avg_pub_c <- subdat |> filter(location == "Central Long Beach") |> na.omit() |> nrow() / central_total
avg_pub_c <- percentage(x = avg_pub_c)

#average public assistance income per neighborhood
public_e <- subdat |> filter(location == "East Long Beach") |> na.omit() |> pull(public_assistance_income) |> mean()
public_s <- subdat |> filter(location == "South Long Beach") |> na.omit() |> pull(public_assistance_income) |> mean()
public_n <- subdat |> filter(location == "North Long Beach") |> na.omit() |> pull(public_assistance_income) |> mean()
public_c <- subdat |> filter(location == "Central Long Beach") |> na.omit() |> pull(public_assistance_income) |> mean()
```

# Data Visualization

Create some visualizations of your data using the esquisse app or the ggplot2 package.

```{r}
#number of surveyed individuals per lb neighborhood in 2023
total_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(east_total, south_total, north_total, central_total))

ggplot(total_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = seq(0, 1500, by = 300)) + xlab("Location") + ggtitle("Number of Surveyed Long Beach Residents in 2023") + ylab("Amount of People") + theme(legend.position = "none") + theme_classic()

#percentage has health insurance per neighborhood
health_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(health_e, health_s, health_n, health_c))

ggplot(health_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + coord_cartesian(ylim = c(50, 100)) + xlab("Location") + ggtitle("Long Beach Health Insurance Coverage") + scale_y_continuous(breaks = seq(50, 100, by = 5), labels = c("50%", "", "60%", "", "70%", "", "80%", "", "90%", "", "100%")) + ylab("Percent") + theme(legend.position = "none") + theme_classic()

#percentage has medicaid per neighborhood
aid_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(med_e, med_s, med_n, med_c))

ggplot(aid_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::label_percent(scale = 1)) + xlab("Location") + ggtitle("Long Beach Medicaid Coverage") + ylab("Percent") + theme(legend.position = "none") + theme_classic()

#percentage has disability per neighborhood
able_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(able_e, able_s, able_n, able_c))

ggplot(able_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::label_percent(scale = 1)) + xlab("Location") + ylab("Percent") + ggtitle("Surveyed LB Residents with a Disability") + theme(legend.position = "none") + theme_classic()

#percent is on food stamps per neighborhood
food_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(food_e, food_s, food_n, food_c))

ggplot(food_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::label_percent(scale = 1)) + xlab("Location") + ylab("Percent") + ggtitle("Surveyed LB Residents on Food Stamps") + theme(legend.position = "none") + theme_classic()

#percentage people have public assistance income in long beach per neighborhood
pub_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(avg_pub_e, avg_pub_s, avg_pub_n, avg_pub_c))

ggplot(pub_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::label_percent(scale = 1)) + xlab("Location") + ylab("Percent") + ggtitle("Surveyed LB Residents with Public Assistance Income") + theme(legend.position = "none") + theme_classic()

#average public assistance income per neighborhood
pub_avg_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(public_e, public_s, public_n, public_c))

ggplot(pub_avg_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = label_dollar()) + xlab("Location") + ylab("Income") + ggtitle("Average Public Assitance Income") + theme(legend.position = "none") + theme_classic()

#average personal income per neighborhood
money_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(money_e, money_s, money_n, money_c))

ggplot(money_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = label_dollar()) + xlab("Location") + ylab("Income") + ggtitle("Average Personal Income") + theme(legend.position = "none") + theme_classic()

#average household income per neighborhood
house_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(house_e, house_s, house_n, house_c)) 

ggplot(house_data, aes(x = Name, y = Value, fill = Name)) + geom_col(stat = "identity", width = 0.5) + scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = label_dollar()) + xlab("Location") + ylab("Income") + ggtitle("Average Household Income") + theme(legend.position = "none") + theme_classic()
```
# Please include additional sections as desired.

```{r}
#tried a bar plot for min, mean, max of household income
east_min <- subdat |> filter(location == "East Long Beach") |> pull(household_income) |> min()
east_max <- subdat |> filter(location == "East Long Beach") |> pull(household_income) |> max()
south_min <- subdat |> filter(location == "South Long Beach") |> pull(household_income) |> min()
south_max <- subdat |> filter(location == "South Long Beach") |> pull(household_income) |> max()
north_min <- subdat |> filter(location == "North Long Beach") |> pull(household_income) |> min()
north_max <- subdat |> filter(location == "North Long Beach") |> pull(household_income) |> max()
cen_min <- subdat |> filter(location == "Central Long Beach") |> pull(household_income) |> min()
cen_max <- subdat |> filter(location == "Central Long Beach") |> pull(household_income) |> max()


plot_data <- data.frame(Name = c("East", "South", "North", "Central"), Value = c(east_min, east_max, house_e, south_min, south_max, house_s, north_min, north_max, house_n, cen_min, cen_max, house_c ))
ggplot(data = plot_data, aes(x = Name, y = Value)) + geom_point(aes(colour = factor(Name))) + scale_y_continuous(breaks = seq(-70000, 1000000, by = 50000), labels = c("-70,000", "", "30,000", "", "130,000", "", "230,000", "", "330,000", "", "430,000", "", "530,000", "", "630,000", "", "730,000", "", "830,000", "", "930,000", "")) + xlab("Location") + ylab("Income") + ggtitle("Min, Mean, and Max Household Incomes") + theme_classic() + theme(legend.position = "none")


```


# Versions

Include some information about the version of R and the packages you are using by running sessionInfo().

```{r}
sessionInfo()
```

Render your document to an html or pdf using the "Knit" button.
